---
title: "Causal Diagrams for Applied Economics"
subtitle: "Energy • Environment • Food Examples (EEFE 530)"
author: "Daniel Brent"
date: "2026-01-21"
format:
  revealjs:
    slide-number: true
    chalkboard: false
    preview-links: true
    incremental: true
    center: false
    theme: [default]
    transition: fade
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## Roadmap (≈1.5 hours)

- Why causal diagrams (DAGs) matter for applied work
- How to *draw* credible DAGs from a research question
- How to read paths: back doors, front doors, colliders, and adjustment
- When to isolate front doors (and when not to)

::: notes
These slides mirror the *structure* of four chapters from *The Effect* (Ch. 6–9), but use different applied economics examples.
:::

---

## Learning objectives

By the end, you should be able to:

- Translate a causal question into a DAG (and defend the assumptions)
- Identify confounding, mediation, and collider bias on a DAG
- Use back-door criteria to propose an adjustment set
- Recognize settings where “front-door-ish” strategies may be useful

---

## Packages and conventions

We’ll use `dagitty` for DAG logic and `ggdag` for plotting.

```{r}
# install.packages(c("dagitty","ggdag","tidyverse"))
library(dagitty)
library(ggdag)
```

Notation:

- Treatment: **T**
- Outcome: **Y**
- Confounder: **C**
- Mediator: **M**
- Unobserved: nodes labeled with `latent = TRUE` in `dagitty`

---

# Chapter 6: Causal Diagrams
## (What a DAG is and what it is not)

---

## 6.1 Causality as intervention

A causal statement is about what would change under an intervention:

- “Increasing **T** would change the distribution of **Y**.”

Applied framing (energy):

- **T:** adoption of time-of-use (TOU) electricity pricing  
- **Y:** peak-period kWh usage

We are not asking: “Are TOU customers different?”  
We are asking: “What would peak usage be *if the same customers* faced TOU?”

---

## 6.1 Causal vs. descriptive language (applied)

In applied papers, watch for “weasel words”:

- “linked to,” “associated with,” “predicts,” “related to”

Example (environment):

- “Counties with more wind farms have lower emissions” (descriptive)
- “Building wind farms reduces emissions” (causal)

When you read: translate to a DAG question.

---

## 6.2 What a causal diagram encodes

A DAG is a compact claim about the **data generating process (DGP)**:

- Nodes = variables we could, in principle, measure
- Directed arrows = direct causal relationships
- No cycles (we’ll discuss later)

DAGs are *models*—they are not “true/false,” but “more/less defensible.”

---

## 6.2 Simple policy effect DAG (food)

Research question: effect of **produce subsidy** on **household fruit/veg intake**

- **T:** exposure to subsidy (eligible + take-up)
- **Y:** intake (servings/day)

```{r}
g1 <- dagitty("dag {
  T [exposure]
  Y [intake]
  T -> Y}")
ggdag(g1, layout = "circle") + theme_dag()
```

---

## 6.2 Adding confounding (energy)

TOU adoption is often **opt-in**.

- **C:** “energy management orientation” (latent) affects adoption and use.

```{r}
g2 <- dagitty("dag {
  T [TOU_adopt]
  Y [peak_kwh]
  C [latent]
  C -> T
  C -> Y
  T -> Y}")
ggdag(g2, layout = "circle") + theme_dag()
```

Implication: naïve comparisons mix causal effect with selection.

---

## 6.3 The real world and omission

A DAG is **not** “everything that matters.”

Goal: include variables that create:
- alternate explanations (back doors)
- channels (front doors)
- biases (colliders)

Rule of thumb: if leaving something out changes your identification story, it belongs.

---

## 6.3 “Measurement” vs “causal” nodes

Sometimes we draw:

- **Constructs** (risk preferences, environmental attitudes)
- **Measurements** (survey item, index, proxy)

In DAGs, be explicit about whether a node is the construct or the proxy.

Example: “Environmental concern” vs. “donated to conservation NGO.”

---

## 6.4 Research questions in DAG terms

A useful template:

1. Define the intervention: what is “changing **T**”?
2. Define the estimand: ATE? effect for compliers? short-run?
3. Translate to: **T → Y** (what blocks spurious paths?)

Applied (environment):

- **T:** nitrate regulation stringency  
- **Y:** drinking water nitrate concentration

---

## 6.5 Moderators (heterogeneity)

Moderators are variables that change the size/sign of a causal effect.

Applied (energy):

- TOU effect may differ by **AC ownership**.

You can represent moderation by:
- including the moderator and discussing heterogeneous effects
- or splitting into subgroup DAGs (same structure, different parameters)

---

# Chapter 7: Drawing Causal Diagrams
## (From a question to a defensible DAG)

---

## 7.1 Start with your idea of the world

Causal diagrams formalize what you already believe.

For applied economics, that belief comes from:

- institutional detail (policy rules, market design)
- engineering/biophysical constraints
- prior evidence
- field knowledge (how people respond)

If you can’t defend arrows, you can’t defend identification.

---

## 7.2 Walk through the DGP (case study)

Case: **Effect of irrigation efficiency upgrades on groundwater pumping**

- **T:** upgrade to drip irrigation
- **Y:** groundwater pumping (acre-feet)

Candidate drivers:

- **Crop mix** (perennials vs annuals)
- **Water prices / pumping costs**
- **Weather / drought**
- **Farm management quality** (latent)
- **Policy incentives / cost-share eligibility**

---

## 7.2 First-pass DAG (include the obvious)

```{r}
g3 <- dagitty("dag {
  T [drip_upgrade]
  Y [gw_pumping]
  W [weather]
  P [pumping_cost]
  K [crop_mix]
  Q [latent]
  S [subsidy_elig]


  W -> Y
  P -> Y
  K -> Y

  Q -> T
  Q -> K
  Q -> Y

  S -> T
  S -> K

  T -> K
  T -> Y}")
ggdag(g3, layout = "fr") + theme_dag()
```

---

## 7.2 Interpret the arrows (discipline-specific)

- **T → Y:** efficiency changes marginal cost of water services; could reduce or increase pumping (rebound).
- **Q (latent):** better managers adopt early and also pump differently.
- **S → T:** cost-share programs increase adoption.
- **S → K:** eligibility can alter crop choices (e.g., perennial conversion).

This is the point where you should argue *mechanisms*.

---

## 7.3 Simplify without losing the identification story

Simplification strategies:

- combine tightly related variables (e.g., “weather” not 10 climate measures)
- omit variables that only sit on already-blocked paths
- collapse multiple mediators into one “behavioral response” node

But: never simplify away your *hard part* (unmeasured confounding, selection).

---

## 7.4 Avoiding cycles

Economics is full of feedback:

- prices affect adoption; adoption affects prices

DAGs are acyclic, so choose a time ordering:

- Use **lagged** versions (e.g., last year’s prices)
- Use a snapshot causal question: “short-run effect holding market fixed”

If feedback is central, consider dynamic models or sequential DAGs.

---

## 7.5 Getting comfortable with assumptions

Two “honesty checks”:

- **Missing arrow** is an assumption (“no direct effect”)
- **Missing node** is an assumption (“doesn’t matter for identification”)

Practice: pick one missing arrow in your DAG and write a defense:
- institutional restriction, timing, physics, or strong prior evidence.

---

# Chapter 8: Causal Paths and Closing Back Doors
## (Read the DAG like an econometrician)

---

## 8.1 Paths explain correlations

A **path** is any route from **T** to **Y** ignoring arrow direction.

Why it matters:

- every open back-door path is an alternate explanation
- your job is to close the bad ones without closing the good ones

---

## 8.2 Finding all paths (small example)

Example: **effect of solar adoption on monthly electricity bills**

- **T:** rooftop solar adoption
- **Y:** bill amount
- **C:** income
- **R:** roof suitability

```{r}
g4 <- dagitty("dag {
  T [solar_adopt]
  Y [bill]
  C [income]
  R [roof_suit]

  C -> T
  C -> Y
  R -> T
  R -> Y
  T -> Y}")
ggdag(g4, layout = "circle") + theme_dag()
```

Paths from T to Y:
- T → Y (front door)
- T ← C → Y (back door)
- T ← R → Y (back door)

---

## 8.3 Good paths vs bad paths (front vs back doors)

- **Front door:** starts with an arrow out of **T**
- **Back door:** starts with an arrow into **T**

Back doors create confounding. Close them by conditioning on appropriate nodes.

In the solar example, conditioning on {Income, Roof suitability} blocks both back doors.

```{r}
adjustmentSets(g4, exposure = "T", outcome = "Y")
```

---

## 8.4 Open and closed paths (conditioning rules)

A path is **closed** if it includes:

- a **non-collider** you condition on
- or a **collider** you *do not* condition on (colliders block by default)

A path is **open** if:
- every collider is conditioned on (or has a conditioned descendant), and
- no non-collider on the path is conditioned on.

These rules drive common “bad control” mistakes in regressions.

---

## 8.5 Collider bias (food marketing example)

Research question: effect of **soda tax exposure** on **soda purchases**

- **T:** tax exposure
- **Y:** purchases
- **U:** health preferences (latent) affect both Y and gym membership
- **G:** gym membership (collider) also affected by income, which affects T

```{r}
g5 <- dagitty("dag {
  T [soda_tax]
  Y [soda_buy]
  I [income]
  G [gym_member]
  U [latent]

  I -> T
  I -> G
  U -> G
  U -> Y
  T -> Y}")
ggdag(g5, layout = "fr") + theme_dag()
```

If you control for **G** (“active lifestyle”), you *open* T ← I → G ← U → Y.

---

## 8.5 What to control for (a regression reading)

Bad controls often arise from:
- post-treatment variables (mediators)
- colliders (or their descendants)

Before writing a regression, ask:

- Is this control a cause of **T**? (potential confounder)
- Is it caused by **T**? (mediator / post-treatment)
- Is it a collider on some back door?

---

## 8.6 Mediation and post-treatment controls (environment)

Research question: effect of **industrial emissions enforcement** on **asthma ED visits**

- **T:** enforcement intensity
- **M:** ambient PM2.5
- **Y:** asthma visits
- **C:** neighborhood SES (confounds T and Y)

```{r}
g6 <- dagitty("dag {
  T [enforcement]
  M [pm25]
  Y [asthma_ed]
  C [ses]

  C -> T
  C -> Y
  C -> M

  T -> M
  M -> Y
  T -> Y}")
ggdag(g6, layout = "fr") + theme_dag()
```

Controlling for **M** gives a *direct effect* of enforcement not mediated by PM2.5, not the total effect.

---

## 8.7 Using paths to test your diagram (placebo logic)

DAGs imply **conditional independencies**.

Example idea:
- If your DAG claims `Roof suitability ⟂ Soda purchases | Income`, you can test it.

Caveat:
- failures may mean the DAG is wrong, or measurement error, or sample selection
- successes do not prove the DAG is correct

Still useful as a diagnostic and for placebo outcomes/exposures.

---

## 8.8 Summary: back-door identification workflow

1. Draw DAG for the question and context
2. List paths from **T** to **Y**
3. Classify: front doors vs back doors
4. Propose adjustment set that blocks all back doors (avoid colliders/post-treatment)
5. Map to estimator (regression, matching, weighting, etc.)

---

# Chapter 9: Finding Front Doors
## (Isolate “good variation” or estimate arrows)

---

## 9.1 Why “close all back doors” can fail

In applied micro, some confounders are:
- unmeasured (preferences, ability, risk tolerance)
- poorly measured (latent constructs)
- endogenous proxies

Front-door thinking asks:
- Can we isolate variation in **T** that avoids key back doors?
- Or can we estimate pieces of the causal chain credibly?

---

## 9.2 Isolating “cleaner” variation (energy natural experiment)

Example: effect of **electricity price** on **consumption**

- **T:** retail electricity price
- **Y:** household kWh

Challenge: price is endogenous to demand and policy.

“Front-door-ish” design:
- isolate price shocks from **fuel cost pass-through** or **regulated tariff formula**
- focus on within-utility, time variation driven by exogenous input costs

DAG intuition: use only the portion of **T** driven by a node with no back door to **Y**.

---

## 9.3 “What the world can do for us”: design strategies

Common ways to isolate front doors:

- randomized encouragement / audit assignment
- rule-based assignment (eligibility cutoffs, lotteries)
- plausibly exogenous shocks (weather for wind generation; disasters; plant outages)
- institutional formulas (tariff pass-through; cap-and-trade allocations)

Key: argue why the source of variation blocks the problematic back doors.

---

## 9.4 Front-door method (conceptual) in an environment setting

Suppose we want the effect of **diesel generator use** (T) on **asthma visits** (Y),
but location and behavior create unobserved confounding.

We observe a mediator:

- **M:** near-home PM2.5 (from monitors / low-cost sensors)

Front-door conditions (informal):

1. All causal influence of T on Y goes through M (T → M → Y)
2. No unblocked back door from T to M
3. No unblocked back door from M to Y once conditioning on T

---

## 9.4 A front-door-style DAG (illustrative)

```{r}
g7 <- dagitty("dag {
  T [diesel_use]
  M [pm25_near]
  Y [asthma_ed]
  U [latent]

  T -> M
  M -> Y

  U -> T
  U -> Y


}")
ggdag(g7, layout = "fr") + theme_dag()
```

Interpretation: we can’t close U by controls, but we may be able to identify via M if conditions hold.

---

## 9.5 Computing front-door quantities (mechanics)

Even when the *total* effect T → Y is confounded, we can estimate:

- Effect of **T on M**
- Effect of **M on Y**, controlling for **T**
- Combine them to recover the overall effect (front-door formula)

In practice, you still need:
- rich measurement of M
- strong assumptions about excluded pathways
- careful robustness checks

---

## 9.6 What can go wrong (and how to diagnose)

Common threats in front-door-style applications:

- T affects Y through other channels not captured by M
- unmeasured confounding between T and M (e.g., weather, site selection)
- unmeasured confounding between M and Y (e.g., concurrent health shocks)
- measurement error in M (attenuation / bias in nonlinear settings)

Diagnostics:
- multiple mediators / negative controls
- sensitivity analysis
- pre-trends / placebo timing tests when using shocks

---

## 9.7 Practical takeaway for applied papers

Most applied economics work uses:

- **back-door closure** (controls, FE, matching)
- or **front-door isolation** via research design (DiD, IV, RD, lotteries)

Use DAGs to:

- justify the identifying variation
- clarify what must be true for identification
- communicate why certain controls are excluded (post-treatment/colliders)
- anticipate critiques (and preempt them)

---

## Wrap-up

What you should do *this week* in EEFE 530:

- Take your term project question and draft a 1-page DAG memo:
  - nodes + arrows + 3 key assumptions
  - identify back doors and propose an adjustment strategy
  - list two plausible threats and one diagnostic per threat

Recommended reference: *The Effect* Ch. 6–9. citeturn1view0turn1view1turn3view0turn3view1
