---
title: "Lecture 04: Potential Outcomes"
subtitle: "EEFE 530"
author: "Daniel Brent"
format:
  revealjs:
    css: ../theme/my-css.css
    theme: [default, ../theme/_extensions/grantmcdermott/clean/clean.scss]
    slide-number: true
    incremental: false
    hash: false
    history: false
    embed-resources: true
    menu:
      side: right
      width: wide
      titleSelector: "h1, h2"
      hideMissingTitles: true
      useTextContentForMissingTitles: false
execute:
  echo: true
  freeze: auto
  warning: false
  message: false
---

```{r, setup, include = F}
# devtools::install_github("dill/emoGG")
library(pacman)
p_load(
  broom, tidyverse,
  latex2exp, ggplot2, ggthemes, ggforce, viridis, extrafont, gridExtra,
  kableExtra, snakecase, janitor,
  data.table, dplyr, estimatr,
  lubridate, knitr, parallel,
  lfe,
  here, magrittr
)
# Define dblue color
dblue <- "#0073e6"
turquoise <- "#20B2AA"
red <- "#CD1818"
red <- "#F92515"
blue <- "#2b59c3"
green <- "#8bb174"
grey_light <- "grey70"
grey_mid <- "grey50"
grey_dark <- "grey20"
green <- "#006400"
slate <- "#314f4f"
# Dark slate grey: #314f4f
# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  warning = F,
  message = F
)
opts_chunk$set(dev = "svg")
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
})
options(crayon.enabled = F)
options(knitr.table.format = "html")
# A blank theme for ggplot
theme_empty <- theme_bw() + theme(
  line = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  axis.text = element_blank(),
  plot.title = element_blank(),
  axis.title = element_blank(),
  plot.margin = structure(c(0, 0, -0.5, -1), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_simple <- theme_bw() + theme(
  line = element_blank(),
  panel.grid = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  axis.text.x = element_text(size = 18, family = "STIXGeneral"),
  axis.text.y = element_blank(),
  axis.ticks = element_blank(),
  plot.title = element_blank(),
  axis.title = element_blank(),
  # plot.margin = structure(c(0, 0, -1, -1), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_axes_math <- theme_void() + theme(
  text = element_text(family = "MathJax_Math"),
  axis.title = element_text(size = 22),
  axis.title.x = element_text(hjust = .95, margin = margin(0.15, 0, 0, 0, unit = "lines")),
  axis.title.y = element_text(vjust = .95, margin = margin(0, 0.15, 0, 0, unit = "lines")),
  axis.line = element_line(
    color = "grey70",
    size = 0.25,
    arrow = arrow(angle = 30, length = unit(0.15, "inches")
  )),
  plot.margin = structure(c(1, 0, 1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_axes_serif <- theme_void() + theme(
  text = element_text(family = "MathJax_Main"),
  axis.title = element_text(size = 22),
  axis.title.x = element_text(hjust = .95, margin = margin(0.15, 0, 0, 0, unit = "lines")),
  axis.title.y = element_text(vjust = .95, margin = margin(0, 0.15, 0, 0, unit = "lines")),
  axis.line = element_line(
    color = "grey70",
    size = 0.25,
    arrow = arrow(angle = 30, length = unit(0.15, "inches")
  )),
  plot.margin = structure(c(1, 0, 1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_axes <- theme_void() + theme(
  text = element_text(family = "Fira Sans Book"),
  axis.title = element_text(size = 18),
  axis.title.x = element_text(hjust = .95, margin = margin(0.15, 0, 0, 0, unit = "lines")),
  axis.title.y = element_text(vjust = .95, margin = margin(0, 0.15, 0, 0, unit = "lines")),
  axis.line = element_line(
    color = grey_light,
    size = 0.25,
    arrow = arrow(angle = 30, length = unit(0.15, "inches")
  )),
  plot.margin = structure(c(1, 0, 1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_set(theme_gray(base_size = 20))
# Column names for regression results
reg_columns <- c("Term", "Est.", "S.E.", "t stat.", "p-Value")
# Function for formatting p values
format_pvi <- function(pv) {
  return(ifelse(
    pv < 0.0001,
    "<0.0001",
    round(pv, 4) %>% format(scientific = F)
  ))
}
format_pv <- function(pvs) lapply(X = pvs, FUN = format_pvi) %>% unlist()
# Tidy regression results table
tidy_table <- function(x, terms, highlight_row = 1, highlight_color = "black", highlight_bold = T, digits = c(NA, 3, 3, 2, 5), title = NULL) {
  x %>%
    tidy() %>%
    select(1:5) %>%
    mutate(
      term = terms,
      p.value = p.value %>% format_pv()
    ) %>%
    kable(
      col.names = reg_columns,
      escape = F,
      digits = digits,
      caption = title
    ) %>%
    kable_styling(font_size = 20) %>%
    row_spec(1:nrow(tidy(x)), background = "white") %>%
    row_spec(highlight_row, bold = highlight_bold, color = highlight_color)
}
```


# Schedule 

## Roadmap

Today’s lecture:

- Define causal effects using the potential outcomes framework
- Introduce counterfactuals and the fundamental problem of causal inference
- Define ATE, ATT, and ATU
- Understand selection bias and bias decomposition
- Connect identification assumptions to common empirical strategies

## Learning Objectives

By the end of this lecture, you should be able to:

- Define potential outcomes and observed outcomes
- State and interpret ATE, ATT, ATU
- Explain why causal effects are not directly observable
- Decompose selection bias in differences-in-means
- State the conditional independence and overlap assumptions
- Interpret regression and matching as identification strategies

## Reading for next week
### Potential outcomes
- [Mixtape Ch. 4]{.orange}
- [Effect Ch. 10]{.dblue}
- [*MHE* Ch. 2]{.red}

## Reading for next week
### Matching
- [Mixtape Ch. 5]{.orange}
- [Effect Ch. 14]{.dblue}
- [*MHE* Ch. 3.2]{.red}
- Matching sections in:
  - Imbens & Woolridge (2009)
  - Caliendo & Kopeinig (2008)
- Applications/Methods Software
  - Diamond and Sekhon (2013)
  - Sekhon (2011); Ferraro & Mirando (2017)
  - Imbens (2015)


# Review 
## Research fundamentals

Angrist and Pischke (MHE) provide four [fundamental questions for research:]{.hi-slate}

1. What is the [causal relationship of interest]{.hi}?

2. How would an [ideal experiment]{.hi} capture this causal effect of interest?

3. What is your [identification strategy]{.hi}?

4. What is your [mode of inference]{.hi}?

::: {.fragment}
Seemingly straightforward questions can be fundamentally unanswerable.
:::

---

# The experimental ideal

## Why are experiments ideal?

Science widely regards [randomized experiments]{.hi} as the [gold standard]{.hi .yellow} for research.

*But why?* The costs can be substantial.

**Costs**

- slow and expensive
- heavily regulated by review boards
- can abstract away from the actual question/setting

**Benefits**

So the benefits need to be pretty large, right?

## Example: Hospitals and health

Imagine we want to know the **causal effect of hospitals on health**.

::: {.fragment}
[Research question]{.hi-slate}

Within the population of poor, elderly individuals, does visiting the emergency room for primary care improve health?
:::

::: {.fragment}
**Empirical exercise**

1. Collect data on *health status* and *hospital visits*.
1. Summarize health status by hospital-visit group.
:::

---

## Example: Hospitals and health (*MHE*)

*MHE* reports empirical exercise from the 2005 National Health Inteview Survey:
```{r, table1, echo = F}
data.frame(
  v1 = c("Hospital", "No hospital"),
  v2 = c("7,774", "90,049"),
  v3 = c(3.21, 3.93),
  v4 = c(0.014, 0.003)
) %>% kable(
  col.names = c("Group", "Sample Size", "Mean Health Status", "Std. Error"),
  align = c("l", "c", "c", "c")
) %>%
kable_styling(font_size = 26) %>%
row_spec(1:2, background = "white", color = slate)
```

::: {.fragment .smaller}
We get a $t$ statistic of 58.9 when testing a difference in groups' means (0.72).
:::


::: {.fragment .smaller}
**Conclusion?** Hospitals make sick people sicker.
:::


::: {.fragment .smaller}
**Alternative conclusion:** Perhaps we're making a mistake in our analysis...
:::


::: {.fragment .smaller}
maybe sick people go to hospitals?
:::


::: {.fragment .smaller}
Any other other other explanations?
:::


## Potential outcomes framework

Let's develop a framework to better discuss the problem here.

::: {.fragment}
- Binary treatment variable (_e.g._, hospitalized): $\text{D}_i = {0,1}$
- Outcome for individual $i$ (_e.g._, health): $\text{Y}_i$

This framework has a few names...

- Neyman potential outcomes framework
- Rubin causal model
- Neyman-Rubin potential outcome/causal  framework/model
:::


## Potential outcomes framework

**Research question:** Does $\text{D}_i$ affect $\text{Y}_i$?

::: {.fragment}
For each individual $i$, there are two [potential outcomes]{.b} (w/ binary $\text{D}_i$)
:::

::: {.fragment}
1. $\color{#0073e6}{\text{Y}_{1i}}$ [if]{.dblue} $\color{#0073e6}{\text{D}_i = 1}$ 
<br> $\color{#0073e6}{i}$['s health outcome if she went to the hospital]{.dblue}
:::

::: {.fragment}
2. $\color{#006400}{\text{Y}_{0i}}$ [if]{.green} $\color{#006400 }{\text{D}_i = 0}$
<br> $\color{#006400}{i}$['s health outcome if she did not go to the hospital]{.green}
:::


::: {.fragment .smaller}
The difference between these two outcomes gives us the [causal effect of hospital treatment]{.hi-red}, _i.e._,

$$
\begin{align}
  \color{#CD1818}{\tau_i} = \color{#0073e6}{\text{Y}_{1i}} - \color{#006400}{\text{Y}_{0i}}
\end{align}
$$
:::

::: {.fragment .smaller}
**Q:** What is $\color{#CD1818}{\tau_i}$?
:::

## Fundamental problem
This simple equation: $\color{#CD1818}{\tau_i} = \color{#0073e6}{\text{Y}_{1i}} - \color{#006400}{\text{Y}_{0i}}$

leads us to [*the fundamental problem of causal inference.*]{.hi-slate}

::: {.fragment .smaller}
> We can never simultaneously observe $\color{#0073e6}{\text{Y}_{1i}}$ and $\color{#006400}{\text{Y}_{0i}}$.  

The same individual cannot simultaneously **go** to the hospital and **not go** to the hospital. Most of applied econometrics focuses on addressing this simple problem.
:::


::: {.fragment .smaller}
Accordingly, our methods try to address the related question
> For each $\color{#0073e6}{\text{Y}_{1i}}$, what is a (reasonably) good counterfactual?

:::


::: {.fragment .smaller}
What determines *"(reasonably) good"* depends on our assumptions.  This is where comparing to the experimental ideal is helpful.
:::

## Solutions?

**Problem:** We cannot directly calculate $\color{#CD1818}{\tau_i} = \color{#0073e6}{\text{Y}_{1i}} - \color{#006400}{\text{Y}_{0i}}$.

::: {.fragment}
**Proposed solution:**
Compare [outcomes for people who visited the hospital]{.dblue} $\left( \color{#0073e6}{\text{Y}_{1i}\mid \color{#0073e6}{\text{D}_{i}=1}} \right)$
<br>to [outcomes for people who did not visit the hospital]{.green} $\left( \color{#006400}{\text{Y}_{0j}\mid \color{#006400}{\text{D}_{j}=0}} \right)$.
:::


::: {.fragment}
$$
\begin{align}
  \mathop{E}\left[ \text{Y}_{i} \mid \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \text{Y}_{i}\mid \color{#006400}{\text{D}_{i}=0} \right]
\end{align}
$$
which gives us the *observed difference in average health outcomes*.
:::


::: {.fragment}
**Q:** This comparison will return *an* answer, but is it *the* answer we want?
:::

## Selection {#selection}

**Q:** What does $\mathop{E}\left[ \text{Y}_{i} \mid \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \text{Y}_{i}\mid \color{#006400}{\text{D}_{i}=0} \right]$ actually tell us?

::: {.fragment}

**A:** A helpful decomposition of  $\text{Y}_{i}$ is
$$
\begin{align}
  \text{Y}_{i}
  &= \color{#006400}{\text{Y}_{0i}} + \text{D}_{i} \underbrace{\left( \color{#0073e6}{\text{Y}_{1i}} - \color{#006400}{\text{Y}_{0i}} \right)}_\color{#CD1818}{\tau_i}
\end{align}
$$
:::


::: {.fragment}

Now write out our expectation, apply this definition, do creative math.

$\mathop{E}\left[ \text{Y}_{i} \mid \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \text{Y}_{i}\mid \color{#006400}{\text{D}_{i}=0} \right]$

:::


::: {.fragment}

<br>  $= \mathop{E}\left[ \color{#006400}{\text{Y}_{0i}} + \text{D}_{i} \left( \color{#0073e6}{\text{Y}_{1i}} - \color{#006400}{\text{Y}_{0i}} \right) \mid \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \color{#006400}{\text{Y}_{0i}} + \text{D}_{i} \left( \color{#0073e6}{\text{Y}_{1i}} - \color{#006400}{\text{Y}_{0i}} \right) \mid \color{#006400}{\text{D}_{i}=0} \right]$

:::


::: {.fragment}

<br>  $= \mathop{E}\left[ \color{#0073e6}{\text{Y}_{1i}}\mid \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \color{#006400}{\text{Y}_{0i}}\mid \color{#0073e6}{\text{D}_{i}=1} \right] + \mathop{E}\left[ \color{#006400}{\text{Y}_{0i}}\mid \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \color{#006400}{\text{Y}_{0i}}\mid \color{#006400}{\text{D}_{i}=0} \right]$

:::

## Selection

**Q:** What does $\mathop{E}\left[ \text{Y}_{i} \mid \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \text{Y}_{i}\mid \color{#006400}{\text{D}_{i}=0} \right]$ actually tell us?

**A:** A helpful decomposition of  $\text{Y}_{i}$ is
$$
\begin{align}
  \text{Y}_{i}
  &= \color{#006400}{\text{Y}_{0i}} + \text{D}_{i} \underbrace{\left( \color{#0073e6}{\text{Y}_{1i}} - \color{#006400}{\text{Y}_{0i}} \right)}_\color{#CD1818}{\tau_i}
\end{align}
$$

Now write out our expectation, apply this definition, do creative math.

$\mathop{E}\left[ \text{Y}_{i} \mid \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \text{Y}_{i}\mid \color{#006400}{\text{D}_{i}=0} \right]$
 <br>  $= \mathop{E}\left[ \color{#006400}{\text{Y}_{0i}} + \text{D}_{i} \left( \color{#0073e6}{\text{Y}_{1i}} - \color{#006400}{\text{Y}_{0i}} \right) \mid \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \color{#006400}{\text{Y}_{0i}} + \text{D}_{i} \left( \color{#0073e6}{\text{Y}_{1i}} - \color{#006400}{\text{Y}_{0i}} \right) \mid \color{#006400}{\text{D}_{i}=0} \right]$ 
<br>  $= \underbrace{\mathop{E}\left[ \color{#0073e6}{\text{Y}_{1i}}\mid \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \color{#006400}{\text{Y}_{0i}}\mid \color{#0073e6}{\text{D}_{i}=1} \right]}_\text{Average treatment effect on the treated 😀} + \mathop{E}\left[ \color{#006400}{\text{Y}_{0i}}\mid \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \color{#006400}{\text{Y}_{0i}}\mid \color{#006400}{\text{D}_{i}=0} \right]$


## Selection

**Q:** What does $\mathop{E}\left[ \text{Y}_{i} \mid \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \text{Y}_{i}\mid \color{#006400}{\text{D}_{i}=0} \right]$ actually tell us?

**A:** A helpful decomposition of  $\text{Y}_{i}$ is
$$
\begin{align}
  \text{Y}_{i}
  &= \color{#006400}{\text{Y}_{0i}} + \text{D}_{i} \underbrace{\left( \color{#0073e6}{\text{Y}_{1i}} - \color{#006400}{\text{Y}_{0i}} \right)}_\color{#CD1818}{\tau_i}
\end{align}
$$

Now write out our expectation, apply this definition, do creative math.

$\mathop{E}\left[ \text{Y}_{i} \mid \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \text{Y}_{i}\mid \color{#006400}{\text{D}_{i}=0} \right]$
 <br>  $= \mathop{E}\left[ \color{#006400}{\text{Y}_{0i}} + \text{D}_{i} \left( \color{#0073e6}{\text{Y}_{1i}} - \color{#006400}{\text{Y}_{0i}} \right) \mid \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \color{#006400}{\text{Y}_{0i}} + \text{D}_{i} \left( \color{#0073e6}{\text{Y}_{1i}} - \color{#006400}{\text{Y}_{0i}} \right) \mid \color{#006400}{\text{D}_{i}=0} \right]$ 
<br>  $= \underbrace{\mathop{E}\left[ \color{#0073e6}{\text{Y}_{1i}}\mid \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \color{#006400}{\text{Y}_{0i}}\mid \color{#0073e6}{\text{D}_{i}=1} \right]}_\text{Average treatment effect on the treated 😀} + \underbrace{\mathop{E}\left[ \color{#006400}{\text{Y}_{0i}}\mid \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \color{#006400}{\text{Y}_{0i}}\mid \color{#006400}{\text{D}_{i}=0} \right]}_\text{Selection bias 😞}$

## Selection

::: { .smallest}
The **first term** is *good variation*—essentially the answer that we want. 

$\mathop{E}\left[ \color{#0073e6}{\text{Y}_{1i}}\mid \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \color{#006400}{\text{Y}_{0i}}\mid \color{#0073e6}{\text{D}_{i}=1} \right]$
:::

::: {.fragment .smallest}
$=\mathop{E}\left[ \color{#0073e6}{\text{Y}_{1i}} - \color{#006400}{\text{Y}_{0i}} \mid \color{#0073e6}{\text{D}_{i}=1} \right]$
:::

::: {.fragment .smallest}
$=\mathop{E}\left[ \color{#CD1818}{\tau_i} \mid \color{#0073e6}{\text{D}_{i}=1} \right]$
:::

::: {.fragment .smallest}
The [average causal effect]{.red} of hospitalization *for hospitalized individuals* -  [Average Treatment Effect on the Treated (ATT)]{.red}.
:::

::: {.fragment .smallest}
The **second term** is *bad variation*—preventing us from knowing the answer.
$\mathop{E}\left[ \color{#006400}{\text{Y}_{0i}}\mid \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \color{#006400}{\text{Y}_{0i}}\mid \color{#006400}{\text{D}_{i}=0} \right]$
:::

::: {.fragment .smallest}
**Selection bias:** The extent to which the "control group" provides a bad counterfactual for the treated individuals.
:::

::: {.fragment .smallest}
**Q:** Can you describe the selection bias $\left( \mathop{E}\left[ \color{#006400}{\text{Y}_{0i}}\mid \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \color{#006400}{\text{Y}_{0i}}\mid \color{#006400}{\text{D}_{i}=0} \right] \right)$ in our hospitalization example in words?
:::


## Selection

Angrist and Pischke (MHE, p. 15):

> The goal of most empirical economic research is to overcome selection bias, and therefore to say something about the causal effect of a variable like $\text{D}_{i}$.

::: {.fragment}
**Q:** So how do experiments—the gold standard of empirical economic (and scientific) research—accomplish this goal and overcome selection bias?
:::

## Back to experiments

::: {.smallest}

**Q:** How do experiments overcome selection bias?

::: {.fragment}

**A:** Experiments break the link between potential outcomes and treatment.
*In other words:* Randomly assigning $\text{D}_{i}$ makes $\text{D}_{i}$ independent of which outcome we observe (meaning $\color{#0073e6}{\text{Y}_{1i}}$ or $\color{#006400}{\text{Y}_{0i}}$).
:::

::: {.fragment}
**Difference in means** with random assignment of $\text{D}_{i}$

$\mathop{E}\left[ \text{Y}_{i} \mid \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \text{Y}_{i}\mid \color{#006400}{\text{D}_{i}=0} \right]$
:::

::: {.fragment}
$= \mathop{E}\left[ \color{#0073e6}{\text{Y}_{1i}}\mid \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \color{#006400}{\text{Y}_{0i}}\mid \color{#006400}{\text{D}_{i}=0} \right]$
:::


::: {.fragment}
$= \mathop{E}\left[ \color{#0073e6}{\text{Y}_{1i}}\mid \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \color{#006400}{\text{Y}_{0i}}\mid \color{#0073e6}{\text{D}_{i}=1} \right]$ from random assignment of ${\text{D}_{i}}$

:::

::: {.fragment}
$= \mathop{E}\left[ \color{#0073e6}{\text{Y}_{1i}} - \color{#006400}{\text{Y}_{0i}}\mid \color{#0073e6}{\text{D}_{i}=1} \right]$
:::

::: {.fragment}
$= \mathop{E}\left[ \color{#CD1818}{\tau_i}\mid \color{#0073e6}{\text{D}_{i}=1} \right]$
:::


::: {.fragment}
$= \mathop{E}\left[ \color{#CD1818}{\tau_i} \right]$
:::


::: {.fragment}
Random assignment of $\text{D}_{i}$ breaks selection bias.
:::

:::

## Randomly assigned treatment

The key to avoiding selection bias: **random assignment of treatment**

::: {.fragment}
<br>(or *as-good-as random assignment*, _e.g._, natural experiments).

:::


::: {.fragment}
Random assignment of treatment gives us
$$
\begin{align}
   \mathop{E}\left[ \color{#006400}{\text{Y}_{0i}}\mid \color{#006400}{\text{D}_{i}=0} \right] = \mathop{E}\left[ \color{#006400}{\text{Y}_{0i}} \mid \color{#0073e6}{\text{D}_{i}=1} \right]
\end{align}
$$
meaning the control group's mean now provides a good counterfactual for the treatment group's mean.
:::


::: {.fragment}
In other words, there is no selection bias, _i.e._,
<br><center>Selection bias $=\mathop{E}\left[ \color{#006400}{\text{Y}_{0i}} \mid \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \color{#006400}{\text{Y}_{0i}}\mid \color{#006400}{\text{D}_{i}=0} \right] = 0$</center>
:::

## Randomly assigned treatment

Additional benefit of randomization:

The [**average treatment effect (ATE)**]{.red} is now representative of the *population average*, rather than the *treatment-group average (ATT)*.

::: {.fragment}
$$
\begin{align}
  \mathop{E}\left[ \color{#CD1818}{\tau_i} \mid \color{#0073e6}{\text{D}_{i}=1} \right] = \mathop{E}\left[ \color{#CD1818}{\tau_i} \mid \color{#006400}{\text{D}_{i}=0} \right] = \mathop{E}\left[ \color{#CD1818}{\tau_i} \right]
\end{align}
$$
:::

## Average Treatment Effect on the Untreated (ATU)

The **Average Treatment Effect on the Untreated (ATU)** answers:

> *What would have happened to those who did **not** receive treatment if they **had** been treated?*

Formally:

$$\text{ATU} \equiv \mathbb{E}[Y_i(1) - Y_i(0) \mid D_i = 0]$$

:::{.smaller}
- $Y_i(1)$: potential outcome **with** treatment  
- $Y_i(0)$: potential outcome **without** treatment  
- $D_i = 0$: individuals who **did not** receive treatment

::: {.fragment}
ATU is a **counterfactual effect for non-participants**.
:::

:::

## Interpreting ATU

ATU is useful when asking **policy-relevant expansion questions**, such as:

- What would happen if we **expanded eligibility**?
- What are the gains from **inducing non-participants** to participate?
- How would outcomes change under **universal treatment**?

::: {.fragment}
ATU generally differs from ATT when **selection into treatment is non-random**.
:::

::: {.fragment}
If treatment effects are heterogeneous:
$$\text{ATE} \neq \text{ATT} \neq \text{ATU}$$
:::

## Comparing ATE, ATT, and ATU

| Estimand | Definition | Interpretation |
|--------|-----------|----------------|
| **ATE** | $\mathbb{E}[Y_i(1) - Y_i(0)]$ | Average effect in the full population |
| **ATT** | $\mathbb{E}[Y_i(1) - Y_i(0) \mid D_i = 1]$ | Effect for those who **received** treatment |
| **ATU** | $\mathbb{E}[Y_i(1) - Y_i(0) \mid D_i = 0]$ | Effect for those who **did not receive** treatment |

::: {.fragment}
These estimands coincide **only if treatment effects are homogeneous**.
:::

::: {.fragment}
With selection on gains:
$$\text{ATE} \neq \text{ATT} \neq \text{ATU}$$
:::


## Graphical Intuition: Sorting on Gains

Think of individuals as having **different treatment gains**:

$$\tau_i = Y_i(1) - Y_i(0)$$

::: {.fragment}
- Individuals with **high gains** are more likely to select into treatment  
- Individuals with **low or negative gains** remain untreated
:::

::: {.fragment}
As a result:
- **ATT** averages gains among *high-$\tau_i$* individuals  
- **ATU** averages gains among *low-$\tau_i$* individuals
:::

::: {.fragment}
Selection bias arises because treatment status reveals information about potential outcomes.
:::

## Graphical intuition: Selection on gains via a latent trait

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 11
#| fig-height: 6
#| fig-cap: "Selection on gains with heterogeneous treatment effects τ(u). Individuals with low latent resistance u select into treatment (u ≤ u*), so ATT averages higher gains than ATU."

library(ggplot2)
library(dplyr)
library(scales)

set.seed(123)

# --- Parameters you can tweak ---
n        <- 600
u_star   <- 0.58         # cutoff u* (treated if u <= u*)
tau0     <- 1.2          # baseline gain
beta     <- -1.8         # slope: negative => higher u => lower gains
noise_sd <- 0.06         # visual texture; set 0 for perfectly smooth

# Generate latent trait u and heterogeneous treatment effects tau(u)
df <- tibble(u = seq(0, 1, length.out = n)) %>%
  mutate(
    tau = tau0 + beta * (u - 0.5) + rnorm(n, 0, noise_sd),
    D   = as.integer(u <= u_star)
  )

att <- df %>% filter(D == 1) %>% summarise(m = mean(tau)) %>% pull(m)
atu <- df %>% filter(D == 0) %>% summarise(m = mean(tau)) %>% pull(m)
ate <- df %>% summarise(m = mean(tau)) %>% pull(m)

# A smoother curve for aesthetics
smooth_df <- df %>%
  mutate(tau_smooth = as.numeric(stats::filter(tau, rep(1/15, 15), sides = 2))) %>%
  tidyr::drop_na()

# Colors (colorblind-friendly-ish)
col_treated   <- "#2C7FB8"   # blue
col_untreated <- "#F03B20"   # orange/red
col_curve     <- "#54278F"   # purple
col_cut       <- "#111111"   # near-black

y_min <- min(df$tau, na.rm = TRUE)
y_max <- max(df$tau, na.rm = TRUE)

ggplot(df, aes(x = u, y = tau)) +
  # Shaded regions (treated vs untreated) with subtle gradients via alpha + border
  annotate(
    "rect",
    xmin = 0, xmax = u_star, ymin = -Inf, ymax = Inf,
    fill = col_treated, alpha = 0.10
  ) +
  annotate(
    "rect",
    xmin = u_star, xmax = 1, ymin = -Inf, ymax = Inf,
    fill = col_untreated, alpha = 0.08
  ) +

  # Light points for texture (optional but helps convey "individual heterogeneity")
  geom_point(color = "grey30", alpha = 0.18, size = 1.1) +

  # Smooth treatment-effect function
  geom_line(
    data = smooth_df, aes(y = tau_smooth),
    linewidth = 1.35, color = col_curve, lineend = "round"
  ) +

  # Cutoff u*
  geom_vline(xintercept = u_star, linetype = "longdash", linewidth = 1.0, color = col_cut) +

  # Mean lines for ATT/ATU/ATE
  geom_hline(yintercept = att, linewidth = 1.0, color = col_treated) +
  geom_hline(yintercept = atu, linewidth = 1.0, color = col_untreated) +
  geom_hline(yintercept = ate, linewidth = 0.9, linetype = "dotted", color = "grey25") +

  # Labels
  annotate(
    "label",
    x = u_star/2, y = att,
    hjust = 1.3, vjust = 0,
    label = paste0("ATT = E[τ(u) | D=1] ≈ ", round(att, 2)),
    fill = alpha(col_treated, 0.12), color = "black",
    size = 4.0
  ) +
  annotate(
    "label",
    x = (1 + u_star)/2, y = atu,
    hjust = -.1, vjust = 0,
    label = paste0("ATU = E[τ(u) | D=0] ≈ ", round(atu, 2)),
    fill = alpha(col_untreated, 0.10), color = "black",
    size = 4.0
  ) +
  annotate(
    "label",
    x = 0.98, y = ate,
    label = paste0("ATE ≈ ", round(ate, 2)),
    hjust = 1, vjust = 0,
    fill = alpha("white", 0.75), color = "black",
    size = 3.8
  ) +
  annotate(
    "text",
    x = u_star, y = y_max,
    label = expression(u^"*"),
    vjust = -0.8, size = 5.2, fontface = "bold"
  ) +

  # Region captions at the bottom
  annotate(
    "text",
    x = u_star/2, y = y_min,
    label = "Treated (D=1): low resistance",
    vjust = 0.0, size = 4.2, color = col_treated, fontface = "bold"
  ) +
  annotate(
    "text",
    x = (1 + u_star)/2, y = y_min,
    label = "Untreated (D=0): high resistance",
    vjust = 0.0, size = 4.2, color = col_untreated, fontface = "bold"
  ) +

  scale_x_continuous(
    labels = percent_format(accuracy = 1),
    breaks = seq(0, 1, by = 0.2)
  ) +
  labs(
    x = "Latent resistance / cost u  (higher = more resistant)",
    y = expression("Treatment gain " * tau(u) * " = Y(1) - Y(0)"),
    title = "Selection on gains (sorting on a latent trait)",
    # subtitle = "Treatment occurs for u ≤ u*. Heterogeneous gains imply ATT and ATU average different parts of τ(u)."
  ) +
  coord_cartesian(ylim = c(y_min, y_max)) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = alpha("grey40", 0.15)),
    panel.grid.major.y = element_line(color = alpha("grey40", 0.15)),
    axis.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey25"),
    plot.margin = margin(10, 10, 10, 10)
  )

```

::: {.fragment}
Key point: if gains 𝜏(𝑢) vary with 𝑢 and selection depends on 𝑢u, then ATT ≠ ATU (sorting on gains).
:::

## Example: Energy efficiency programs

Governments subsidize energy efficiency investments (sometimes specifically targeting low-income homeowners)

::: {.fragment}
**Q:** Do these programs have the desired effects (_i.e._, lower energy use)?
:::

::: {.fragment}
**A:** Observational studies—comparing energy data from participants and non-participants—may not be a good approach].
:::

::: {.fragment}
**Challenges:** Participants self select. **+** Programs target lower-income homeowners.
:::


## Example: Energy Efficiency programs

How do we formalize these concerns in our framework?

::: {.fragment .smaller}
**Observational program evaluations:**
<br>
$\quad\mathop{E}\left[ \text{Energy}_{i}\mid \color{#0073e6}{\text{Program}_{i}=1} \right] - \mathop{E}\left[ \text{Energy}_i \mid \color{#006400}{\text{Program}_{i}=0} \right]=$
$$
\begin{align}
&\underbrace{\mathop{E}\left[ \color{#0073e6}{\text{Energy}_{1i}}\mid \color{#0073e6}{\text{Program}_{i}=1} \right] - \mathop{E}\left[ \color{#006400}{\text{Energy}_{0i}}\mid \color{#0073e6}{\text{Program}_{i}=1} \right]}_{\text{Average causal effect of energy efficiency program on energy use for participants, }\mathit{i.e.} \text{, } \overline{\tau}_{1}} \, + \\[0.5em]
&\underbrace{\mathop{E}\left[ \color{#006400}{\text{Energy}_{0i}}\mid \color{#0073e6}{\text{Program}_{i}=1} \right] - \mathop{E}\left[ \color{#006400}{\text{Energy}_{0i}}\mid \color{#006400}{\text{Program}_{i}=0} \right]}_\text{Selection bias}
\end{align}
$$
:::

::: {.fragment .smaller}
If the program attracts/selects individuals who, on average, have higher energy use without the program (sort of the point of the program), then we have positive selection bias (decreasing estimated effect of the ATT in absolute value).
:::


::: {.fragment}
**Q:** Is this the only type of selection bias we should expect for this type of program?
:::

## Example: Energy Efficiency programs

$\quad\mathop{E}\left[ \text{Energy}_{i}\mid \color{#0073e6}{\text{Program}_{i}=1} \right] - \mathop{E}\left[ \text{Energy}_i \mid \color{#006400}{\text{Program}_{i}=0} \right] =$
$$
\begin{align}
&\mathop{E}\left[ \color{#0073e6}{\text{Energy}_{1i}}\mid \color{#0073e6}{\text{Program}_{i}=1} \right] - \mathop{E}\left[ \color{#006400}{\text{Energy}_{0i}}\mid \color{#0073e6}{\text{Program}_{i}=1} \right] \, + \\[0.5em]
&\mathop{E}\left[ \color{#006400}{\text{Energy}_{0i}}\mid \color{#0073e6}{\text{Program}_{i}=1} \right] - \mathop{E}\left[ \color{#006400}{\text{Energy}_{0i}}\mid \color{#006400}{\text{Program}_{i}=0} \right] \\[0.5em]
\end{align}
$$

So even if the program, on average, has an negtive impact on energy use (in the participant group), _i.e._, $\overline{\tau}_1>0$, we will underestimate the magnitude if there is positive selection bias.

::: {.fragment}
If the bias is sufficiently large (relative to the treatment effect), our estimate may even get the sign of the effect wrong.
:::


::: {.fragment}
**More to come:** We will go through an example in water conservation that highlights these challenges: [Brent & Ward (2018)](https://www.sciencedirect.com/science/article/pii/S0095069619300014)
:::


# The experimental ideal
## Example: The STAR experiment [(*MHE*)]{.orange}

The Tennessee STAR experiment is a famous/popular example of an experiment that allows us to answer an important social/policy question.

**Research question:** Does classroom size (or resources more generally) affect student performance?

::: {.fragment}
- Statewide(-ish) in Tennessee for the 1985–1986 kindergarten cohort
- Ran for 4 years with ∼11,600 children. Cost ∼$12 million.
:::


::: {.fragment}
**Treatments:**

  1. *Small* classes (13–17 students)
  1. *Regular* classes (22–35 students) plus part-time teacher's aide
  1. *Regular* classes (22–35 students) plus full-time teacher's aide
  
:::

---

**First question:** Did the randomization balance participants' characteristics across the treatment groups?

::: {.fragment}
Ideally, we would have pre-experiment data on the outcome variable (test scores).

Unfortunately, we only have a few demographic attributes.
:::

---

```{r, table221, echo = F, escape = F}
tab221 <- data.frame(
  v1 = c("Free lunch", "White/Asian", "Age in 1985", "Attrition rate",
    "K. class size", "K. test percentile"),
  v2 = c(0.47, 0.68, 5.44, 0.49, 15.10, 54.70),
  v3 = c(0.48, 0.67, 5.43, 0.52, 22.40, 48.90),
  v4 = c(0.50, 0.66, 5.42, 0.53, 22.80, 50.00),
  v5 = c(0.09, 0.26, 0.32, 0.02, 0.00, 0.00)
) %>% kable(
  escape = F,
  digits = 2,
  col.names = c("Variable", "Small", "Regular", "Regular + Aide", "P-value"),
  align = c("l", rep("c", 4)),
  caption = "Table 2.2.1, MHE"
) %>%
row_spec(1:6, background = "white", color = slate) %>%
column_spec(1, color = "black", italic = T) %>%
add_header_above(c(" " = 1, "Treatment: Class Size" = 3, " " = 1))
tab221
```

---

```{r, table221_2, echo = F, escape = F}
tab221 %>% row_spec(1:3, color = dblue, bold = T)
```

Demographics appear balanced across the three treatment groups.

::: {.fragment}
Maybe not for free lunch, but the differences are pretty small.
:::

---

```{r, table221_3, echo = F, escape = F}
tab221 %>% row_spec(4, color = dblue, bold = T)
```

The three groups differ significantly on attrition rate.

---

```{r, table221_4, echo = F, escape = F}
tab221 %>% row_spec(5, color = dblue, bold = T)
```

The randomization generated variation in the treatment.

---

```{r, table221_5, echo = F, escape = F}
tab221 %>% row_spec(6, color = dblue, bold = T)
```

The small-class treatment significantly increased test scores.

# The experimental ideal
## The STAR experiment

The previous table estimated/compared the treatment effects using simple differences in means.

We can make the same comparisons using regressions.

Specifically, we regress our outcome (test percentile) on dummy variables (binary indicator variables) for each treatment group.

## Regression analysis

Assume for the moment that the treatment effect is constant<sup>[†]{.dblue}</sup>, _i.e._,

$$
\begin{align}
  \color{#0073e6}{\text{Y}_{1i}} - \color{#006400}{\text{Y}_{0i}} = \color{#CD1818}{\rho} \quad \forall i
\end{align}
$$

::: {.fragment}
then we can rewrite
$$
\begin{align}
  \text{Y}_{i} = \color{#006400}{\text{Y}_{0i}} + \text{D}_{i}\left( \color{#0073e6}{\text{Y}_{1i}} - \color{#006400}{\text{Y}_{0i}} \right)
\end{align}
$$
:::


::: {.fragment}
as
$$
\begin{align}
  \text{Y}_{i} = \underbrace{\alpha}_{=\mathop{E}\left[ \color{#006400}{\text{Y}_{0i}} \right]} + \text{D}_{i} \underbrace{\color{#CD1818}{\rho}}_{\color{#0073e6}{\text{Y}_{1i}} - \color{#006400}{\text{Y}_{0i}}} + \underbrace{\eta_i}_{\color{#006400}{\text{Y}_{0i}} - \mathop{E}\left[ \color{#006400}{\text{Y}_{0i}} \right]}
\end{align}
$$

:::

::: {.footnote}
[†]{.dblue}You'll often hear econometricians say "homogeneous" (*vs.* "heterogeneous").
:::

## Regression analysis

::: {.smallest}
$$
\begin{align}
  \text{Y}_{i} = \alpha + \text{D}_{i} \color{#CD1818}{\rho} + \eta_i
\end{align}
$$

Now write out the conditional expectation of $\text{Y}_{i}$ for both levels of $\text{D}_{i}$
:::

::: {.fragment .smallest}
$\mathop{E}\left[ \text{Y}_{i} \mid \color{#0073e6}{\text{D}_{i}=1} \right] =$
:::

::: {.fragment .smallest}
$\mathop{E}\left[ \alpha + \color{#CD1818}{\rho} + \eta_i \mid \color{#0073e6}{\text{D}_{i}=1} \right]$
:::

::: {.fragment .smallest}
$=\alpha + \color{#CD1818}{\rho} + \mathop{E}\left[ \eta_i | \color{#0073e6}{\text{D}_{i}=1} \right]$
:::

::: {.fragment .smallest}
$\mathop{E}\left[ \text{Y}_{i} \mid \color{#006400}{\text{D}_{i}=0} \right]$
:::

::: {.fragment .smallest}
$= \mathop{E}\left[ \alpha + \eta_i \mid \color{#006400}{\text{D}_{i}=0} \right]$
:::

::: {.fragment .smallest}
$= \alpha + \mathop{E}\left[ \eta_i \mid \color{#006400}{\text{D}_{i}=0} \right]$
:::

::: {.fragment .smallest}
Take the difference...
$\mathop{E}\left[ \text{Y}_{i} \mid \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \text{Y}_{i} \mid \color{#006400}{\text{D}_{i}=0} \right]$
:::

::: {.fragment .smallest}
$= \color{#CD1818}{\rho} + \underbrace{\mathop{E}\left[ \eta_i | \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \eta_i \mid \color{#006400}{\text{D}_{i}=0} \right]}_\text{Selection bias}$
:::

## Regression analysis

$$
\begin{align}
  \mathop{E}\left[ \text{Y}_{i} \mid \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \text{Y}_{i} \mid \color{#006400}{\text{D}_{i}=0} \right] = \color{#CD1818}{\rho} + \mathop{E}\left[ \eta_i | \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \eta_i \mid \color{#006400}{\text{D}_{i}=0} \right]
\end{align}
$$

Again, our estimate of the [**treatment effect**]{.red} $\left( \color{#CD1818}{\rho} \right)$ is only going to be as good as our ability to eliminate the **selection bias**.

**Selection bias in regression model:** $\mathop{E}\left[ \eta_i | \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \eta_i \mid \color{#006400}{\text{D}_{i}=0} \right]$

Selection bias here should remind you a lot of

::: {.fragment}
**omitted-variable bias.**
There is something in our disturbance $\eta_i$ that is affecting $\text{Y}_{i}$ and is also correlated with $\text{D}_{i}$.
:::

::: {.fragment}
In other econometrics parlance: Our treatment $\text{D}_{i}$ is *endogenous* if it is correlated with the error term $\eta_i$.
:::

## Solutions and covariates {#covariates}

**Selection bias in regression model:** $\mathop{E}\left[ \eta_i | \color{#0073e6}{\text{D}_{i}=1} \right] - \mathop{E}\left[ \eta_i \mid \color{#006400}{\text{D}_{i}=0} \right]$

::: {.fragment}
As before, if we randomly assign $\text{D}_{i}$, then selection bias disappears.
:::

::: {.fragment}
Another potential route to identification is to condition on covariates in the hopes that they "take care of" the relationship between $\text{D}_{i}$ and whatever is in our disturbance $\eta_i$.
:::

::: {.fragment}
Without very clear reasons explaining how you've controlled for **all** the "bad variation", clean and convincing identification on this path is going to be challenging.
:::

## Covariates

That said, covariates can help with two things:

1. Even experiments may need **conditioning/controls:** The STAR experiment was random *within school*—not across schools.

1. Covariates can reduce residual variance—[increasing precision.]{.hi-slate}

::: {.fragment}
Now that we've seen regression can analyze experiments, let's estimate the STAR example...
:::

---

```{r, table222, echo = F, escape = F}
tab222 <- data.frame(
  v1 = c("Small class", "", "Regular + aide", "", "White/Asian", "", "Female", "",
    "Free lunch", "", "School F.E."),
  v2 = rbind(
    c(4.82, 0.12, "", "", ""),
    c("(2.19)", "(2.23)", "", "", "")
  ) %>% as.vector() %>% c("F"),
  v3 = rbind(
    c(5.37, 0.29, "", "", ""),
    c("(1.26)", "(1.13)", "", "", "")
  ) %>% as.vector() %>% c("T"),
  v4 = rbind(
    c(5.36, 0.53, 8.35, 4.48, -13.15),
    c("(1.21)", "(1.09)", "(1.35)", "(0.63)", "(0.77)")
  ) %>% as.vector() %>% c("T")
) %>% kable(
  escape = F,
  col.names = c("Explanatory variable", "1", "2", "3"),
  align = c("l", rep("c", 4)),
  caption = "Table 2.2.2, MHE"
) %>%
row_spec(1:11, color = slate) %>%
row_spec(seq(2,11,2), color = "#c2bebe") %>%
row_spec(1:11, extra_css = "line-height: 110%;") %>%
column_spec(1, color = "black", italic = T)
tab222
```

The omitted level is *Regular* (with part-time aide).

---


```{r, table222_2, echo = F}
tab222 %>% column_spec(2, bold = T)
```

Results without other controls are very similar to the difference in means.

---

```{r, table222_3, echo = F}
tab222 %>% column_spec(3, bold = T)
```

School FEs enforce the experiment's design and increase precision.

---

```{r, table222_4, echo = F}
tab222 %>% column_spec(4, bold = T)
```

Additional controls slightly increase precision.

# Future work
## Future work
The rest of the course will focus on methods and designs to eliminate selection bias and estimate causal effects.  

We will try to think about these in the context of the causal framework.

Some of these that you should be comfortable implementing (by the end of the course) are:

- Instrumental variables
- Matching
- Difference-in-difference designs
- Synthetic control
- Regression discontinuity


